Uncertainty in social learning evolution

### 2022-05-18 sci.tex
started modifying figures, reviewing existing model features in table in terms of our model factors

Social learning is essential to survival. It is likely to evolve when it is more
efficient than asocial, trial-and-error learning. The consensus in cultural
evolutionary theory holds that some amount of environmental variability and
uncertainty about the best decisions are necessary for social learning to evolve.
However, current models for the evolution of social learning tend to conflate forms
of uncertainty, and rarely consider different ones in tandem. Moreover, many models
are limited by considering only two possible behaviors and environmental states.
Here we use evolutionary agent-based modeling to improve on these shortcomings. We
model a time-varying environment with dozens of possible behaviors performed by
agents engaging in individual and social learning. We show that ambiguous payoffs,
larger possible decision sets, and shorter agent lifespans sometimes increase social
learning prevalence, as expected. However, we also find that, under some conditions,
these forms of uncertainty can select against social learning.
\end{abstract}

\section{Introduction}

\subsection{The history of "copy when uncertain"}
It has become a near truism that "theory predicts that it is adaptive to learn when uncertain."
This claim has been motivated by empirical work (such as Asch conformism studies) \& theoretical models (Boyd Richerson 85, 88) that often do not mention the word uncertainty at all. However, they do show related effects of social learning increasing when individual learning is difficult (i.e., ambiguous lines in Asch), and intermediate levels of temporal uncertainty favoring social learning (i.e., too stable and you get genetic canalization of a trait, too much temporal variation and there's selection for individual learning to ensure the most up to date information).

{Galef 2009} https://www.sciencedirect.com/science/article/pii/S006534540939004X\#bb0059
cites {B \& R 85/88 \& Laland 2004} https://link.springer.com/content/pdf/10.3758/BF03196002.pdf
Laland 2004 cites {B\& R 88}
\url{https://books.google.com/books?hl=en&lr=&id=mqFcAgAAQBAJ&oi=fnd&pg=PA29&dq=Boyd+and+Richerson\%E2\%80\%99s+(1988&ots=IfIbB_kx_a&sig=g8CYH5bVfwfXhTRGot0q761f9mw\#v=onepage&q&f=false}

Other examples of empirical work: https://www.sciencedirect.com/science/article/pii/S0022096516300625

Henrich \& Boyd 88 do have something like B that is the difficulty of getting right answer (or rho probability of doing so)
https://www.sciencedirect.com/science/article/pii/S109051389800018X\#FIG5

Lifespan not modelled

\subsection{earlier structure}

Social learning is essential to human and other species' everyday life and survival.
It allows individuals to solve problems when acquiring information from others is more efficient than learning
on one's own~\cite{Laland2004}. Theory predicts that social
learning should be favored in contexts with greater
uncertainty~\cite{BoydRicherson1985,Henrich1998}, and this prediction has received some empirical support~\cite{McElreath2005,Kendal2018}. 
However, the meaning of the term ``uncertainty'' is not always clear, and often conflates environmental variability, spatial heterogeneity, and ambiguity or uncertainty about payoff structure. 
Moreover, most models of the evolution of social learning ``blackbox'' key 
cognitive learning processes that underlie it~\cite{Heyes2016}.

In this paper we use agent-based modeling to compare the effect of
different sources of uncertainty on social learning
by ``unblackboxing'' typically abstracted-out model components 
of environmental variability, payoff structures and
agent life histories, and  learning mechanisms. ``Uncertainty''  
means variability where the probabilistic structure is unknown. %cannot be learned.
Uncertainty increases when payoffs are more similar across behaviors, when
environmental variability increases, when the number of possible behaviors
increases, and when lifespan decreases.  In this paper we show that more ambiguous
payoff structures and shorter lifespans sometimes do lead to greater reliance on
social learning---however, we also identify and explain cases where greater
uncertainty leads to less social learning due to the possibility that social
information is misleading.  We thus conclude that many predictions made by previous
models of the evolution of social learning are likely overgeneralized.
%~\cite{Yarkoni2021}.

\subsection{Social Learning}

Social learning, as we consider it here, occurs whenever an individual acquires a behavior by observing another individual. This need not require explicit instruction, and is, in fact, widespread across a broad range of nonhuman 
taxa~\cite{Kendal2018,Allen2019}. Importantly, social information can be inherited both
from parents --- i.e., via \emph{vertical transmission} like genetic information --- and
from others in the same generation --- i.e., via \emph{horizontal transmission}
\cite{CavalliFeldman1981}. The joint action of vertical and horizontal transmission
gives rise to qualitatively different evolutionary dynamics. For example,
inter-generational environmental change will affect the adaptive value of genetic
information and vertically-transmitted cultural information more than information
that is horizontally transmitted. We include both horizontal and vertical transmission pathways in our
model. For simplicity, we ignore oblique transmission in which non-parental members
of the previous generation are observed.

Environmental variability has been seen as a key selective force in shaping social learning starting with the first formal models of cultural evolution~\cite{CavalliFeldman1981,BoydRicherson1985}.
Totally stable environments will not favor learning mechanisms because information can become genetically hardwired, while  extreme environmental instability will degrade the value
of social learning as information becomes rapidly outdated~\cite{Feldman1996}. This suggests that an intermediate degree of environmental predictability will favor social learning. Strategies can also evolve to mitigate the risks of relying on outdated social information by weighing more heavily information from others who more recently acquired it~\cite{Rendell2010}.

Uncertainty has also been modelled as arising from other aspects of the environment. 
For example, \citeA{Perreault2012} vary the ambiguity of the environmental cue individuals get through individual learning about the state of the world. Perhaps not surprisingly, the more ambiguous the asocial information, the greater the selection for weighing social information heavily. Alternatively, uncertainty about the optimal behavior has been modelled by increasing the number of cultural traits to choose from.

Empirical research supports some of these theoretical predictions. Organisms flexibly use social learning as a function of the ambiguity of the environmental cue and of other environmental features that are often subsumed under the rubric of ``uncertainty.''
While some studies explicitly impose a cost whenever participants use asocial
information \cite{Morgan2012,Atkisson2012}, 
others allow the costs of each strategy to emerge as a function of task structure and assess its consequences for learning strategies. 
For example, when participants received equivocal private information about the best investment to make in a lab game,
they were more likely to rely on social information to make their choice~\cite{Toelch2013}. 
\citeA{McElreath2005} developed a similar experiment where participants ``pulled'' virtual slot machine arms (often called ``bandits''), each yielding stochastic payoffs. Participants relied more on social learning when the bandits had higher-variance payoffs, and when the highest-paying bandit changed more frequently. 
The number of options to choose between can also increase uncertainty about the optimal choice, and has been shown to increase participants' reliance on social learning \cite{Muthukrishna2016a}.

Thus existing theoretical work and the empirical evidence seem to support that various forms of uncertainty favor the evolution of social learning. However, uncertain outcomes are operationalized in different ways across models, and any given model tends to focus on only one or two forms of uncertainty at a time. Our agent-based modeling approach enables us to explicitly specify different forms of uncertainty independently in order to understand which of these environmental factors particularly favor the evolution of social learning. Simultaneous modelling also allows us to examine their interaction. We first attempt conceptual replications of previous models' findings, and then examine where they diverge.

\subsection{Research overview}

Computational agents in our model face a problem: every time step they perform
one of several behaviors, with each behavior represented by a ``bandit'' that
pays off 1 or 0 with some probability. One of these behaviors (the optimal
behavior) pays off with a higher probability than all the others.
Agents decide which behavior to 
perform based either on a success-biased observation 
of a peer's behavior (social learning) or based on their own observations (asocial learning).
Agents then update their memory of expected payoffs for each behavior 
when they receive a payoff from their chosen
action. Within this framework we have four mechanisms by which we operationalize 
and vary uncertainty:
(1) the expected payoffs of the optimal behavior and all the rest---when payoffs
are nearly identical, uncertainty in the form of ambiguity increases; (2) the
environmental variability, i.e., the probability that the optimal behavior 
changes between generations; (3) the number of possible behaviors the environment allows---which
behavior is optimal is more uncertain when there are more possibilities; and (4)
 agent lifespan---agents experience fewer learning opportunities and die more uncertain about which behavior
is optimal when their lifespan is shorter.

The primary outcome measure of our
model is the average difference between the frequency of (horizontal) 
social learning and the frequency of asocial
learning across all agents. If social learning is more prevalent than asocial
learning this suggests that the optimal behavior is more likely found
by copying peers than by trial-and-error search.
Conversely, when asocial learning is more prevalent, this suggests social information
is likely to be misleading. 
When social and asocial learning are 
equally prevalent this means there is no discernible advantage to either, i.e.,
the agents have weak priors on which channel provides more reliable information. 
Each agent has its own social learning frequency that it inherits from its
parent (haploid reproduction) with mutation, so evolution selects for,
or computes~\cite{Smaldino2013}, the optimal
social learning frequency.

Using our model we found that increased uncertainty sometimes
led to increased reliance on social learning, as expected from prior literature. However, we also find cases where increased uncertainty decreased agents' reliance on social learning
due to increased uncertainty that made social information less reliable, thereby
increasing reliance on asocial learning.

\section{Model}

We developed an agent-based model of a society of $N$ individuals who each must decide
which of $B$ behaviors to perform at each time step. Each behavior is a ``bandit'',
a common modelling and experimental approach for representing behaviors with
probabilistic payoffs~\cite{SuttonBartoBook, McElreath2005,Rendell2010,Schulz2019}. %Bandits are like  slot machines in a casino~\ref{fig:behaviorSelection}.
Each behavior $b$ yields Bernoulli payoffs: payoff of 1 with probability $\pi_b$ and zero payoff
with probability $1 - \pi_b$; 
$\pi_b$ is therefore also the expected payoff of behavior $b$. Agents must
decide which behavior to perform at each time step. To do this, agents 
use an explore-exploit strategy to sometimes try the most profitable behavior
they know about, and other times try alternatives that may pay off more reliably.

We operationalized uncertainty in four different ways: 
(1) payoff ambiguity, $A_\pi$, which measures the difference
between the optimal expected payoff behavior $\pisub{high}$ and the expected payoff
of the other behaviors, $\pisub{low}$; 
(2) environmental variability, $u$, the probability the optimal behavior changes from one generation to another; 
(3) the number of possible behaviors, $B$; and 
(4) the lifespan, or time steps per generation, $L$.

We ran the simulation until agents evolved to be all social or all asocial 
learners, or until 20k time steps was reached. Only 44 out of 396k trials
did not reach fixation (all social or all asocial). At
the end of each generation agents reproduce and then die off. 
Those selected to reproduce
pass on their boolean social learning trait, $s$, without mutation,
which specifes whether agents learn socially. We developed a series of
computational analyses where we systematically vary the uncertainty
variables and observe the average prevalence of social learning trait $s$
across agents and trials, which is our primary outcome measure, 
denoted $\meansl$.

\subsection{Agents and their attributes}

In each time step, $N$ agents---autonomous problem solvers---select 
which behavior to perform based on their running tally of mean payoffs for 
each available behavior. %information (Figure~\ref{fig:behaviorSelection}). 
Agent $i$ tracks the mean payoff of each behavior $b$, denoted $\bar\pi_{ib}$, and
a \emph{count} of how many times it has performed each behavior, denoted $c_{ib}$.
$\bar\pi_{ib}$
is initialized to 0 for all $b$ at model initialization for
all agents, and after each generation for individual learner agents. Social
learning agents' $\bar\pi_{ib}$ is initialized as that of its teacher selected
through performance-biased oblique learning. Agent $i$'s accumulated payoffs
from performing several behaviors over their lifetime of $L$ time steps is
denoted $\pi_{i}$.

\begin{table}[h]
    \caption{Dynamic agent-level variables. Each has an implicit time dependence.}
    \label{tab:modelParameters}
    \centering %\hspace{-3em}
    \begin{tabular}{cp{2.5in}p{1.0in}} \toprule

        Attribute & Description & Initial value \\

        \midrule

        $s_i$  & Social learner trait: 1 if agent $i$ is social learner; 0 otherwise & 0
        or 1 w/ equal prob. \\

        $\bar\pi_{ib}$ & ``Ledger'': mean payoffs acquired via behavior $b$ by $i$ 
                       & $B$-vector of floating-point zeros \\

        $c_{ib}$ & Count of how many times agent $i$ performed $b$ 
              & $B$-vector of integer zeros \\

        $\pi_i$ & Net payoffs accumulated by $i$ within generation
                                & 0.0 \\
        \bottomrule
    \end{tabular}
\end{table}
\subsection{Modeling uncertainty}

In our model uncertainty is a tunable consequence of four environmental features:
\begin{enumerate}
    \item 
      We vary the frequency at which the optimal behavior changes, which we call
      \emph{environmental variability}, $u$. At the start of each generation, with
      probability $u$, a new behavior is assigned a payoff of $\pisub{high}$ while all
      other behaviors are assigned a payoff of $\pisub{low}$. Otherwise, the same
      behavior remains optimal across generations. 
    \item We vary the latent expected payoffs yielded by the bandits. For simplicity,
      we assume that in any given environmental state, there is one optimal behavior that
      yields an expected payoff of $\pisub{high}$, while all other behaviors yield a 
      payoff of $\pisub{low} < \pisub{high}$. As $\pilow$ increases towards 
      $\pihigh$, uncertainty in the form of \emph{payoff ambiguity} also increases,
      making it more difficult, and less rewarding, for agents to determine which
      behavior is the optimal one.
    \item We vary the total \emph{number of available behaviors}, $B$, which is a
      source of uncertainty since agents are less likely to know which behavior yields high payoffs.  
    \item Finally, we vary number of behavioral events per generation, $L$. This can be viewed as the \emph{effective lifespan} of an agent.  Decreasing this lifespan effectively increases the importance of each event for acquiring payoffs. When lifespans are short ($L$ is small), agents experience greater uncertainty about which behavior is optimal given the fewer learning opportunities.
\end{enumerate}

\subsection{Dynamics and evolution}

Model dynamics can be split into three parts: initialization,
\emph{intra}generational behavior and payoff accumulation, and
\emph{inter}generational transmission of the social learning trait and oblique
learning (Figure~\ref{fig:IntraInterGenerationalDynamics}). 
The model runs that we analyze here were initialized with $N=100$ agents
to randomly have the social learner trait or not, with no accumulated payoffs, and
blank ``ledgers''. At each time step, agents select one of $B$ possible behaviors
to perform, accumulating a payoff of 1 if the bandit pays off, and 0 if it does
not---a bandit pays off with probability $\pihigh$ if it is the unique optimal
behavior, or with probability $\pilow$ if it is one of $B-1$ non-optimal
behaviors. Agents accumulate payoffs in this way for $L$ time steps in each
generation. At the end of each generation, $N=100$ agents are randomly selected with
replacement to reproduce (asexual, haploid reproduction), 
weighted by their net payoffs over the generation,
i.e., \emph{performance-biased} reproduction. Child agents then learn from 
one teacher, selected again via a form of performance bias: a child agent chooses
its teacher by first selecting a fully random subset of agents from the 
parent generation as potential teachers, then selecting the one with the 
greatest net payoffs among the subset. This process continues until agents
have become all social or all asocial learners, or when $20k$ time steps have
elapsed.

\begin{figure}
  \caption{\textbf{Model dynamics.} 
    The first generation of agents are initialized with empty ledgers, 
    then perform behaviors on each subsequent time step (A). At the end
    of each generation, after all agents have performed $L$ behaviors
    chosen using softmax selection, the intergenerational tramsission stage
    begins (B). During this stage agents reproduce, passing on their social
    learning trait without mutation (B1). Child agents learn from
    the previous generation if they are social learners (B2), or are initialized
    with blank ledgers if they are individual learners. All agents from the previous
    generation die off, and new child agents begin within-generation dynamics anew.}
  \label{fig:IntraInterGenerationalDynamics}
\hspace{-2em}
  % \centering 
    \includegraphics[width=1.1\textwidth]{Figures/IntraInterGenerationalDynamics.pdf}
\end{figure}

\subsubsection{Initialization}

Model initialization includes agent and environment initializations. Agents are
randomly initialized to have the social learning trait or not, with equal 
probability. Agent ``ledgers'' and behavior counts 
are uniformly initialized to the no-knowledge,
blank-slate case where all $\pi_{ib} = 0$ and $c_{ib} = 0$. Both the 
ledger and behavior count vectors have $B$ entries, one for each
behavior the environment affords. Net payoffs for
the generation are also initialized to zero. All agents are initialized with
the same softmax temperature, $\tau$, that guides their behavior selection 
to be either more exploratory (greater $\tau$) or more exploitative of
what agents believe to be the optimal behavior(s) (lesser $\tau$).

The environment is initialized to have $B$ behaviors with one of them selected
at random to be the optimal behavior, denoted $b^*$. Behavior $b^*$ is 
initialized to have a probability $\pihigh$ of paying off 1. All other behaviors
are initialized to have probability $\pilow$ of paying off 1. The modeler must
specify the environmental variability, $u$, and the number of time steps per 
generation, $L$.

\vspace{2em}
\begin{table}[h]
    \caption{Global environmental and cognitive initialization parameters.
    \mt{TODO: Test more $N_\text{T}$ for sensitivity analysis}}
    \label{tab:modelParameters}
    \centering %\hspace{-3em}
    \begin{tabular}{cp{2.5in}p{1in}} \toprule

        Symbol & Description & Values tested (bold$=$default) \\

        \midrule

        $B$       & Number of possible behaviors (represented by ``bandits'') 
                  & 2, 4, 10 \\

        $\pihigh$ & Probability that the unique optimal behavior pays off 1 
                & \textbf{0.9} \\

        $\pilow$ & Probability one of $B - 1$ non-optimal behaviors pays off 1 
                 & 0.1, 0.45, 0.8 \\

        $\tau$ & Softmax temp.; $\uparrow=$more exploration, $\downarrow=$more
                    exploitation 
               & 0.01, \textbf{0.1}, 1.0 \\

        $u$    & Probability optimal behavior changes between generations 
               & 0.0, 0.1, \ldots, 1.0 \\

        $L$    & Number of time steps per generation & 1, $B/2$, $B$, $2B$, $4B$ \\

        $N_T$    & Number of teachers to pool, from which best selected 
                 & \textbf{5}  \\

        \bottomrule
        \end{tabular} 
\end{table}

\subsubsection{Intra-generational behaviors and payoffs}

Each generation begins with agents initialized either by the $t=0$ initialization
outlined above, or initialized via inter-generational reproduction and learning,
described in detail below. Within generations, there is no social learning.
At each time step within a generation, agents select a behavior to perform
using the softmax algorithm, then update their ledgers $\bar\pi_{ib}$ and behavior
counts $c_{ib}$. If 
the chosen bandit pays off for an agent, its net payoff is incremented by one,
$\pi'_i \leftarrow \pi_i + 1$. This process continues for $L$ time steps, 
at which point reproduction, learning, and die-off occur, which re-initializes 
the next generation for performing $L$ behaviors selected via this same process.

At each time step, 
agent $i$ chooses behavior $b$ at random, with each behavior
weighted by the softmax function applied to that behavior's observed mean payoff
relative to all mean payoffs in the ledger,
\begin{equation}
  \Pr(i \text{ chooses behavior } b) = \frac{\exp(\bar\pi_{ib}
  / \tau) }{ \sum_{b=1}^B \exp(\bar\pi_{ib} / \tau)}.
\end{equation}  
Softmax behavior selection is a
biologically plausible model of behavior~\cite{Schulz2019} 
that enables agents to explore
alternative behaviors sometimes and exploit the best observed behavior other times,
in accordance with Luce's choice axiom~\cite{Luce1959},.
The parameter $\tau$ specifies how frequently alternative behaviors are
explored versus how frequently the best observed behaviors are 
exploited. Greater $\tau$ means more frequent exploration, lesser $\tau$ means
more frequent exploitation. We set $\tau = 0.1$ for
all computational analyses presented in the main paper. We 
 performed
sensitivity analyses and found a weak dependence on $\tau$ that does not affect our
main conclusions \mt{have run $\tau$ sensitivity checks, need to analyze them}.

When agent $i$ performs behavior $b$, $i$'s behavior count 
is incremented by 1, $c'_{ib} \leftarrow c_{ib} + 1$. Agent $i$'s ledger of mean 
payoffs are updated % from $\bar\pi_{ib}$ to $\bar\pi_{ib}'$
using exponential weighted averaging,

\begin{equation}
  \bar\pi_{ib}' = \bar\pi_{ib} +
    \frac{\mathrm{Bandit_{b}(0, 1)} - \bar\pi_{ib}}{c_{ib}'},
\end{equation}
where 
$\mathrm{Bandit}_{b}(0, 1)$
is 0 or 1 depending on the result of the bandit draw for behavior $b$.

\subsubsection{Inter-generational inheritance and social learning}

Every $L$ time steps, agents from the current generation reproduce, teach,
and die, while the next generation to perform the next $L$ time steps learns
from one teacher from the previous generation if they inherited the social 
learner trait. Inter-generational dynamics thus depend on 
two payoff-biased selection mechanisms: reproducer selection
and teacher selection.

$N$ reproducers are sampled from the population with
replacement weighted by net payoffs, i.e., at each of $N$ draws for a parent,
\begin{equation}
  \Pr(\text{Agent $i$ chosen for reproduction}) = \frac{\pi_i}{\sum_{i=1}^N \pi_i}.
\end{equation}
\noindent
We sample with replacement to allow for more successful agents to have multiple
offspring. Child agents inherit their parents' social learning trait, so 
all social learner parents spawn social learner offspring, and parents lacking
the trait spawn offspring lacking the trait.

Child agents with the social learning trait then must select and learn from a
teacher from their parent's generation, including possibly their parent.
Child agents give no preference to whether potential teachers are social 
learners. A child selects a teacher by first selecting a pool of $N_T = 5$ potential
teachers. Among this pool, each child selects the teacher with the greatest
net payoffs. In case of a tie the teacher is chosen at random from those with
the shared maximum net payoff.

Social learner child agents each acquire their teacher's ledger, and all social
learner children behavior counts are reset to 1 to limit ledger values to be between
0 and 1. Child agents without the social learner trait have all ledger and behavior
count values re-initialized to 0. At this point the re-initialized child agents
engage in behavior selection and payoff accumulation described above.

\subsection{Computational analyses and outcome measures}

We manipulated environmental uncertainty parameters described above, $u$,
$\pisub{low}$, $B$, and $L$, to examine their effects on our
main outcome measure, the mean prevalence of social learning $\meansl$, 
and supplemental outcome measures $\meanpi / L$ and $\meanT / L$. $\meanpi / L$ is the
mean net payoffs accumulated by agents at the end of each generation for the
final generation of agents, normalized by lifespan $L$. $\meanT / L$ is the 
mean time steps to convergence normalized by the lifespan of agents, i.e., the
mean number of generations until convergence was reached. Mean values were
caluculated across 100 model runs for each combination of uncertainty parameter values.
To test the effect of the uncertainty parameters, 
we initialized several model runs with
systematically varied uncertainty parameters, $u$, $\pisub{\mathrm{low}}$, $B$,
and $L$. We varied $u \in \{0.0, 0.1, \ldots, 1.0\}$ for each combination of
the following parameters:
$\pisub{low} \in \{0.1, 0.5, 0.8\}$, $B \in \{2, 4, 10\}$, and
$L \in \{1, B, 2B, 4B\}$ for $B=2$ and $L \in \{1, B/2, B, 2B\}$ for $B=4,10$.

\begin{table}[h]
    \caption{Outcome variables}
    \label{tab:outcomeVariables}
    \centering %\hspace{-3em}
    \begin{tabular}{cp{2.5in}p{1in}} \toprule

        Symbol & Description & Values \\

        \midrule

        $\meansl$ & Mean social learning prevalence over agents and trials
                  & $\in [0.0, 1.0]$ \\

        $\meanpi / L$ & Mean payoffs accumulated in a generation normalized by
        lifespan & $\in [0.0, 1.0]$ \\

        $\meanT / L$ & Mean number of generations to convergence & Bounded above by 20k / 8\\
        \bottomrule
    \end{tabular}
\end{table}

\subsection{A note on relationship with standard RL formalism and models}

We presented our model as an agent-based model because that is
how we originally developed it. Howevever, we believe the model could be 
translated in a straightforward way where each action at time $t$, typically denoted
$a_t$, is to ``pull'' one of the $B$ bandits, and the policy $\pi$ would 
involve using the softmax function to weight random behavior
selection~\cite{SuttonBartoBook}. Because of the complication of two
relevant time intervals, intervals between actions and generational intervals, and
space limitations we do not try to put the entire model in standard RL terms.
\mt{Expand/improve.}

\section{Analysis}

To understand the effect of uncertainty parameters (1) environmental variability, $u$; (2) payoff
ambiguity that increases with $\pilow$; (3) number of behavioral options, $B$; and 
(4) generation length/agent lifespan, $L$, on the evolution of social learning, we analyze the results of
model runs across systematically varied parameter combinations.  We observed complex
outcomes where the effects of uncertainty parameters were highly dependent on one
another as measured by the mean prevalence of social learning, $\meansl$, across
trials for each parameter setting.  Increased environmental variability, $u$, 
tended to suppress social learning as expected from standard models of
social learning evolution. However, the shape of the decrease in $\meansl$ 
was not uniform (Figure~\ref{fig:evolutionOfSL}). 
Depending on the parameter settings, the critical value of $u$ at which social 
learning suppression began varied, as did the steepness of the decrease. 
We found that the location and width of the decrease in $\meansl$ is correlated
with the difference between the expected net payoffs of a population of all
social learners, $\meansoc$, and the expected net payoffs
of an all-asocial learner population, $\meanasoc$ (Figure~\ref{fig:meanNetPayoffs}). 
evolution does not always choose the theoretically highest payoff strategy,
and confirmed that evolution is more uncertain (i.e.\ selection is
weaker~\cite[p. 103]{BoydRicherson1985}) when $\meansoc - \meanasoc$ is small, 
where evolutionary uncertainty is measured by the 
average steps to model convergence, $\meanvar{T}$ (Figure~\ref{fig:steps}).

\subsection{Evolution of social learning under uncertainty}

We observed that different uncertainty contexts lead to different patterns in
the evolution and suppression of social learning. To understand the evolution
and suppression of social learning we inspected the social learning prevalence,
$\meansl$, which is the average of the frequency of social learning across
a model population over all trial model runs for a 4-tuple of uncertainty
values, $(u, \pilow, B, L)$.  We plot $\meansl$ on the y-axis 
over $u=0.0,0.1,\ldots,1.0$ on the x-axis in each plot, with nine plots total over
$\pilow=0.1,0.45,0.8$ and $B=2,4,10$, broken out by $L=1,B/2,B,2B$ within each plot
(Figure~\ref{fig:evolutionOfSL}). Across uncertainty scenarios, $\meansl$ monotonically
decreases $u$ increases, in accordance with classic cultural evolutionary
predictions~\cite{CavalliFeldman1981,BoydRicherson1985,Feldman1996}.  Therefore we compare
differences between uncertainty contexts in terms of the \emph{suppression of social learning}
over (increasing) $u$.

Differences emerged in the suppression of social learning on two dimensions.
First there were systematic differences in the value of $u$ when social
learning first begins to be suppressed, i.e., where $\meansl$ first decreases. We show
in the following subsection that this occurs when $\meansoc$ approaches $\meanasoc$.
The second dimension of analysis is the width of the transition from 
maximal to minimal $\meansl$. In some cases, this width was less than 1, but
in many other cases the width was equal to 1, meaning that $\meansl$ 
decreased over all $u$. This indicated weak selection pressure since 
less than 0.001\% of all trials ended with all agents either social or asocial
learners. This is confirmed below by examining the average number of time steps
it took to reach fixation (agents all social or all asocial).

For now we will examine differences in the location and width of social 
learning suppression. \mt{In outline form while I work out each point}
\begin{itemize}
  \item 
    First, note that as $\pilow$ increases towards $\pihigh = 0.9$, social learning
    suppression widens, becoming linearly decreasing over $u$ with no $\meansl$
    reaching 1 or 0. This is due to weakened selection as the difference between
    social and asocial learning payoffs goes to zero.
  \item
    Second, as $B$ increases, note that social learning suppression 
    begins at increasingly high $u$. This is because it is increasingly worth
    the risk that a socially-learned ledger biases agents towards non-optimal
    behaviors when the number of choices increases. Furthermore, when $B$ is
    larger, one large ledger entry is not as misleading as when $B$ is smaller
    since there is more probability mass in the softmax vector's non-maximal
    behavior entries when $B$ is larger, and so alternatives to the socially
    learned ``best'' behavior will be explored more frequently.
  \item
    Finally, note that decreased lifespan both increases the location of 
    social learning suppression and the width of social learning suppression,
    with the extreme case of $L=1$ becoming linear and eventually flat compared
    to other values of $L$ as $\pilow$ and $B$ increase.
\end{itemize}
To support these explanations we need to analyze the average payoffs 
obtained by agents in these various uncertainty contexts and compare them
with the expected payoffs when all agents are individual learners, and with the
expected payoffs when all agents are social learners.

\subsection{Relative benefit of social learning}

Social learning begins to be suppressed when the expected payoffs from 
a population of all social learners, $\meansoc$ (Figure~\ref{fig:meanNetPayoffs};
small dashed line with diamond markers),
decreased to the point where they are nearly the same as expected payoffs for
a population of all individual learners, $\meanasoc$
(Figure~\ref{fig:meanNetPayoffs}; horizontal long-dash lines). We see that social 
learning becomes completely suppressed when $\meansoc < \meanasoc$.  Mean 
payoffs realized by model agents approximately track expected payoffs of
whichever strategy is optimal, social or asocial learning, but with some important
variation.

Similarly, sometimes populations that start out mixed outperform all
social learner populations when $\meansoc > \meanasoc$, indicating that an initial
diversity of learner types helps with the long-term prosperity of the group, even
though the asocial learner type tends to die out. \mt{I suspect
this may be sensitive to $N_T$, so my next analysis will be the effect
of $N_T$.}

Second, notice that asocial learning may be the safe fallback option when 
evolution is unsure which strategy is optimal, 
but a population of all social learners would do better
than what evolution tended to select for (Figure~\ref{fig:meanNetPayoffs},
e.g., when $\pilow=0.45$, $B=2$, and $L=8$ in the middle left plot). \mt{I suspect
this will also be sensitive to $N_T$.}

\subsection{Selection strength and evolutionary uncertainty}

If selection is truly weaker when $\meansoc \approx \meanasoc$, then we should not
only see a more perfect bimodality of outcomes, with roughly half of outcomes going
to both $\meansl = 1$ and $\meansl = 0$, but we should also see the models taking
longer to converge in this situation.  Indeed, weak selection is also evidenced by
spiking or elevated average time to convergence, $\meanT$, around when theoretical
$\meansoc$ and observed $\meanpi$ approach $\meanasoc$. Recall that models are run
until they reach fixation or until they reach 20k time steps. Only 44 out of 396k
did not reach fixation, less than 0.001\% of all trials.

\begin{figure}
  \caption{Dependence of Social Learning ($\meansl$) on environmental
  variability $u$ for several values of low payoff $\pilow$, number of
  behaviors $B$, and lifespan $L$. Average over 1000 trials.} 
  \label{fig:evolutionOfSL}
  \includegraphics[width=\textwidth]{mainResultsPlots.pdf}

\input{mainResultsPlots.tex}

\input{meanNetPayoffs.tex}

\input{stepResultsPlots.tex}

\section{Discussion}

\begin{itemize}
  \item 
    Increasingly, engineers are blurring the line between humans and AI agents in designing
    effective interventions to the wicked problems we face
    today~\cite{Irrgang2021,Rolnick2022}.
  \item
    We have connected such augmented intelligence technologies as MARL to
    cultural evolution, which is a promising approach to addressing existential
    threats to life on Earth~\cite{Jones2021}.
\end{itemize}

% \section{Model}

% We developed an agent-based model of a society of $N$ individuals who each must decide
% which of $B$ behaviors to perform at each time step. Each behavior is a ``bandit'',
% a common modelling and experimental approach for representing behaviors with
% probabilistic payoffs~\cite{SuttonBartoBook, McElreath2005,Rendell2010,Schulz2019}. %Bandits are like  slot machines in a casino~\ref{fig:behaviorSelection}.
% Each behavior $b$ yields a payoff of 1 with probability $\pi_b$ and a payoff of zero
% otherwise; $\pi_b$ is therefore also the expected payoff of behavior $b$. Agents must
% decide which behavior to perform at each time step. To do this, agents 
% use an explore-exploit strategy to sometimes try the most profitable behavior
% they know about, and other times try alternatives that may pay off more reliably.

% We operationalized uncertainty in four different ways: 
% (1) payoff ambiguity, $A_\pi$, which measures the difference
% between the optimal expected payoff behavior $\pisub{high}$ and the expected payoff
% of the other behaviors, $\pisub{low}$; 
% (2) environmental variability, $u$, the probability the optimal behavior changes from one generation to another; 
% (3) the number of possible behaviors, $B$; and 
% (4) the lifespan, or time steps per generation, $M$.

% We ran the simulation for $T=1000$ time steps. At
% the end of each generation agents reproduce and then die off. 
% Those selected to reproduce
% pass on their boolean social learning trait, $s$, without mutation,
% which specifes whether agents learn socially. We developed a series of
% computational analyses where we systematically vary the uncertainty
% variables and observe the average prevalence of social learning trait $s$
% across agents and trials, which is our primary outcome measure, 
% denoted $\langle s \rangle$.

% \subsection{Agents and their attributes}

% In each time step, $N$ agents---autonomous problem solvers---select 
% which behavior to perform based on their running tally of mean payoffs for 
% each available behavior. %information (Figure~\ref{fig:behaviorSelection}). 
% Agent $i$ tracks the mean payoff of each behavior $b$, denoted $\bar\pi_{ib}$, and
% a \emph{count} of how many times it has performed each behavior, denoted $c_{ib}$.
% $\bar\pi_{ib}$
% is initialized to 0 for all $b$ at model initialization for
% all agents, and after each generation for individual learner agents. Social
% learning agents' $\bar\pi_{ib}$ is initialized as that of its teacher selected
% through performance-biased oblique learning. Agent $i$'s accumulated payoffs
% from performing several behaviors over their lifetime of $M$ time steps is
% denoted $\pi_{i}$.

% \vspace{2em}

% \begin{table}[h]
%     \caption{Dynamic agent-level variables. Each has an implicit time dependence.}
%     \label{tab:modelParameters}
%     \centering \hspace{-3em}
%     \begin{tabular}{cp{4.0in}l} \toprule

%         Attribute & Description & Initial value \\

%         \midrule

%         $s_i$  & Social learner trait: 1 if agent $i$ is social learner; 0 otherwise & 0
%         or 1 w/ equal prob. \\

>>>>

Social learning is essential to survival. It is likely to evolve when it is more
efficient than asocial, trial-and-error learning. Theoretically, some uncertainty is
necessary for social learning to be necessary, however too much uncertainty makes
social information useless as information becomes outdated. This fact is empirically
supported across biology and human sciences. However, we lack a theoretical
framework predict the effects of specific classes and types of uncertainty on social
learning. Furthermore, existing models and experimental operationalizations of
uncertainty are ambiguously related, and models of uncertainty and social learning
tend to only consider a small number of sources of uncertainty and behavioral
choices.  Here we use evolutionary agent-based modeling to improve on these
shortcomings. We model a time-varying environment with a varying number
of possible behaviors agents could perform to acquire payoffs.
We show that ambiguous payoffs, larger possible decision sets, and shorter agent
lifespans interact to affect social learning in complex ways that make evolutionary
selection more or less certain about the optimal strategy. Our work consolidates
models of uncertainty and social learning and models the evolution of social
learning that could be used to guide human, non-human, and artificial life
towards optimal responses to existential threats and new opportunities.
\end{abstract}

Social learning is essential to human and other species' everyday life and survival.
It allows individuals to solve problems when acquiring information from others is
more efficient than learning on one's own~\cite{Laland2004}. Theory predicts that
social learning should be favored in contexts with greater
uncertainty~\cite{BoydRicherson1985,Henrich1998}, and this prediction has received
some empirical support across species~\cite{McElreath2005,Kendal2018,Allen2019}.
However, the meaning of the term \emph{uncertainty} is not always clear, and often
conflates environmental variability, spatial heterogeneity, ambiguity or
uncertainty about payoff structure, and other possible interpretations of
\emph{uncertainty}.  Moreover, most models of the do not account for
individual-level cognition~\cite{Heyes2016}.
